---
title: "AI & LLM Example"
author: "Erik Westlund"
date: "2025-06-11"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"
editor:
  render-on-save: true


---

## Overview

* We'll use an LLM to help us prepare data for a visualization
* Let's use the [CDC's data on influenza vaccination coverage for all ages](https://data.cdc.gov/Flu-Vaccinations/Influenza-Vaccination-Coverage-for-All-Ages-6-Mont/vh55-3he6/about_data)

## Data Preparation

```{r setup}
#| echo: false
#| message: false

# Install required packages if not already installed
required_packages <- c("dplyr", "ggplot2", "forcats", "janitor", "kableExtra",
                      "readr", "stringr", "tidyr")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load required packages
library(dplyr)
library(ggplot2)
library(forcats)
library(janitor)
library(kableExtra)
library(readr)
library(stringr)
library(tidyr)

```

## Overview

* We'll use an LLM to help us prepare data for a visualization
* Let's use the [CDC's data on influenza vaccination coverage for all ages](https://data.cdc.gov/Flu-Vaccinations/Influenza-Vaccination-Coverage-for-All-Ages-6-Mont/vh55-3he6/about_data)

## Data Preparation

```{r}
#| message: false

df <- read_csv(here::here("data", "raw", "cdc_Influenza_Vaccination_Coverage_for_All_Ages__6__Months__20250610.csv"))

df |> glimpse()
```

## Chat with the LLM

### Data Preparation

From here, we can chat with the LLM to help us prepare the data for a visualization.

We'll start with the most lazy prompt we can, and see how it goes.

```
I am completely unqualified for my job. Can you look at the below notebook, clean up the data, and create me a plot of influenze vaccine coverage by age group?

My job depends on it and I only have five minutes, please help. I just saw the boss's car pull into the parking lot.
```
